
## Basis

The span (columnspace) of a matrix $A$ is denoted enoted $C(A)$.
A basis for $C(L)$ is denoted $B_L$ while $w$ is the generic form of vectors in $L$:

```{r}
B_L <- columnspace(L)
v <- vec(B_L, "v")
```

$$ B_L = `r tex(B_L)`, \quad v = `r tex(v)` $$

## Basis for orthogonal complement

A basis for the orthogonal complement $W=C(L)^\perp$ of $L$ can be found as follows:
$W$ consists of the vectors $v$ that are orthogonal to any vector in
$C(L)$; that is to any vector of the form $Lx$. In matrix notation,
$W=\{v|v' L x=0 \quad \forall x\}$. That is $x' L' v=0$. For this to
be satisfied for all $x$ we must have $L'v=0$, that is $v$ is in the
null space of $L'$:

```{r}
B_W <- nullspace(t(L))
w <- vec(B_W, "w")
```

$$
B_W = `r tex(B_W)`, \quad
w = `r tex(w)`
$$




Vectors in $L$ and $V$ are indeed orthogonal:
```{r}
sum(v * w)
```

## Null space

The null space of L1 is spanned by
```{r}
N <- nullspace(t(L1))
b <- vector_sym(ncol(N), "b")
```

$$
N=`r tex(N)`, \quad
b = `r tex(b)`,
n=`r tex(N %*% b)`,
n=`r tex(vec(N, "n"))`
$$




## Intersection of two spaces

Let $U$ and $V$ be bases and matrices. A basis for the intersection can be found as:
We look for $C(U)\cap C(V)$ ie vectors $z$ of the form

$$
	z = Ux = Vy
$$

for $(x,y)$ non-zero. That is, we find the null space of $A=[U | - V]$ because if $A(x,y)´=0$ then $Ux=Vy$. For example

```{r}
U = nullspace(t(L1))
V = L2
A = cbind(U, -V)
N_A <- nullspace(A)
```

$$
A = `r tex(A)`,\quad
N_A = `r tex(N_A)`
$$

So the vectors in the intersection $U\cap V$ have the form:
```{r}
x <- N_A[1:3,]
y <- N_A[4:5,]
U %*% x
V %*% y
z <- vec(U %*% x, "z")
z
```

$$
U = `r tex(U)`; \quad
V = `r tex(V)`; \quad
z = `r tex(z)`
$$

##  projections

```{r}
II <- diag_(1, 4)
P_L1 <- L1 %*% inv(t(L1) %*% L1) %*% t(L1)
R <- II-P_L1
```
